Synthesis of Linear Functional Programs

### 2021-07-20 article/article.tex
tweaks and further todos

% General motivation?
Program synthesis is an automated or semi-automated process of deriving a
program, i.e.~generating code, from a high-level specification. Synthesis can be
seen as a means to improve programmer productivity and program correctness
(e.g. through suggestion and autocompletion).
%
Specifications can take many forms such as natural
language~\cite{chen2021evaluating},
examples~\cite{DBLP:conf/popl/FrankleOWZ16} or rich types such as
polymorphic refinement types~\cite{DBLP:conf/pldi/PolikarpovaKS16} or
graded types~\cite{DBLP:conf/lopstr/HughesO20}. Regardless of the kind
of specification, program synthesisers have to deal with two main
inherent sources of complexity -- searching over a vast space of
(potentially) valid programs, and interpreting user intent.

@@

The focus of our work is on type-driven synthesis where specifications take the
form of linear types.
%
% TODO: Deixar isto para mais à frente?  More concretely, we will explore the
% proof-theoretic foundations of linear types via the Curry-Howard
% correspondence with linear logic propositions, viewing program synthesis as
% proof search in linear logic.  (Said correspondence maps \emph{types} to
% \emph{propositions in a logic}, and \emph{programs} to \emph{proofs}).
%
Linear types differ from more traditional types in that they constrain resource
usage in programs by \emph{statically} limiting the number of times certain
resources can be used during their lifetime (linear resources must be used
\emph{exactly once}).
%
They can be applied to resource-aware programming such as concurrent programming
(e.g. session types for message passing
concurrency~\cite{DBLP:journals/mscs/CairesPT16}), memory-management
(e.g.~Rust's ownership types), safely updating-in-place mutable
structures~\cite{Bernardy_2018}, enforcing protocols for external \textsc{api}s~\cite{Bernardy_2018}, to name a few.
% TODO: Repensar esta parte dos exemplos de aplicação e de como podemos
% relacionar o nosso trabalho com futuras aplicações nestes campos

\paragraph{Contributions.} Despite their long-known potential (\emph{linear
types can change the world!}\cite{Wadler90lineartypes},
\cite{DBLP:journals/mscs/CairesPT16,Bernardy_2018}) and strong proof-theoretic
foundations
\cite{10.1093/logcom/2.3.297,DBLP:conf/cade/ChaudhuriP05,DBLP:journals/tcs/CervesatoHP00},
synthesis with linear types in a more advanced context
has generally been overlooked in the literature.  In this paper we present a framework for synthesis
with linear types extended with recursive algebraic data types, parametric
polymorphism and refinements.
% TODO: extensible? proving feasability of... ? showing potential for ... ?

\paragraph{Outline.} First, we discuss linear types as specifications, the
outline of the synthesis process, and what that entails for the synthetiser --
ilustrated with concrete results (Section~\ref{sec:overview}). Then, we present
the formal system driving the synthesis (Section~\ref{sec:formal_system}).
Aftwerwards, we describe the architecture of our synthetiser named \synname, and
examine technical details and challenges of the implementation
(Section~\ref{sec:architecture}). Finally, we evaluate our work through
expressiveness benchmarks (i.e.~can we synth $X$?) and time benchmarks (i.e.~how
fast can we synth a term of type $T$?) (Section~\ref{sec:evaluation}).
%TODO: And appendice with...

@@

% TODO: Why linear types?
The \synname\ synthetiser combines the features of linear types with recursion,
parametric polymorphism, recursive algebraic data types, and
refinement types. The synthetiser is built on top of a system of proof-search in
linear logic. Proof-search relates to program synthesis through the Curry-Howard
correspondence~\cite{}, which states that \emph{propositions in a logic} map
directly to \emph{types}, and \emph{proofs} of those propositions to well-typed
\emph{programs} -- finding a proof of a proposition is finding a program with
that type.

Linear types make for more precise specifications than simple traditional types
because information on which resources must necessarily be used is encoded in
the type. Their preciseness also influences the search space: all programs where
a linear resource is used non-linearly (i.e. not \emph{exactly once}) are
ill-typed. With linearity built-in to the synthesis process, usage of a
linear proposition more than once is not even considered, and unused ones are caught
mid-process, making proof-search in linear logic potentially faster than
in classical logic.  % TODO: Reword?

The core of the synthesis is a \emph{sound} and \emph{complete} system
consisting of \emph{bottom-up} proof-search in propositional linear logic based
on \emph{focusing}~\cite{}. Our approach, being grounded by
propositions-as-types, ensures that all synthetised programs (i.e.~proofs) are
well-typed \emph{by construction} (i.e.~if the synthesis procedure terminates,
then the program satisfies its specification). Moreover, we can leverage the
modularity of the proof-search based approach in two key axes: first, since
proof search need not construct only closed proofs, we can effectively
synthetise program sub-expressions (i.e.~synthesis based on typed holes);
secondly, the framework is amenable to extensions to the core propositional
language, allowing for the introduction of a richer type structure while
preserving the correctness of programs by construction.

The \synname\ programming language is naturally born alongside the synthetiser
to create an environment for it. Synth goals are inserted in the program by use
of a \emph{synth} keyword or mark, indicating a program should be generated for a given
type. Additionally, it opens the possibility of synthesis with context, that is,
with knowledge of other functions and structures defined in the same program.

\paragraph{Core linear calculus.} Initially, we synthetise from
specifications that only use literals and canonical linear types ($\lolli, \with, \tensor, 1,
\oplus, \bang$), such as linear logic theorems. Namely,
the specification $(A \lolli B \lolli C) \lolli (A \tensor B) \lolli C$ produces
the program \mintinline{haskell}{(\a -> \b -> let c*d = b in a c d)}
(known as \emph{uncurry}).

\paragraph{Beyond propositional logic.} To enhance \synname\ and its language,
counter the core's lack of expressiveness, synthetise more interesting programs
-- and thus, empirically prove the feasability of more relevant synthesis in a
linear context, we extend the syntax and type system with recursiveness,
parametric polymorphism, algebraic data types (ADTs) and type refinements.

@@

\paragraph{ADTs.} We reiterate the expressiveness of linear types: by requiring some types to be
used linearly (consumed), we can assure, e.g., the deconstruction of an
ADT. As such, the specification $\bang(a \lolli b) \lolli List\ a \lolli
List\ b$ isn't satisfied by the program \mintinline{haskell}{(\x -> \y -> Nil)}.
To synthetise a valid program from it, \mintinline{haskell}{(List a)} has to be
deconstructed, and its constructor arguments used. \synname\ would output the following
function (\emph{map}) without relying on any additional information:
% TODO: Aqui não estou bem a falar de polymorphism mas parece estranho não por
% visto que falo dos outros temas? se calhar faço um à parte?
%
\begin{minted}{haskell}
map c d = let !e = c in
    case d of
          Nil  -> Nil
        | Cons f -> let g*h = f in Cons (e g, map (!e) h);
\end{minted}

\paragraph{Refinements.} Refinements are predicates added to a type which must hold for any
element that inhabits said type. The \synname\ language supports simple
refinements on integers with arithmetic expressions in the predicates. They
assert the robustness of our main framework, i.e., how it is amenable to
significant additions without necessarily interfering with the algorithm.
%
As an example, from the specification $x \{Int\} \lolli y \{Int\} \lolli z
\{Int\} \lolli k \{Int\ \vert\ x + k = y * z\}$, the
program\ \mintinline{haskell}{(fn a b c = add (mult (sub 0 1) a) (mult c b))} is
synthetised.

\paragraph{Configuration.} Specifications can be additionally augmented with three keywords: ``using'', ``depth'' and
``assert''. Through them, we can fine-tune the synthesis process and, respectively,
force certain functions to be present in the synthesis outcome, allow a
\emph{deeper} search in the valid programs space, and require given predicates
to hold true in the program after the synthesis is complete.

Taken directly from the Linear Haskell paper~\cite{Bernardy_2018}, we present a
complex example of synthesis: given the linear type signatures for array
primitives (\emph{newMArray} passes a new mutable array to a function that uses
it linearly, \emph{write} takes a mutable array and writes a pair to it,
\emph{freeze} consumes a mutable array and produces an imutable array,
\emph{foldl} has the default definition) we try to synthetise a function
``array'' that provides an immutable array given a list of pairs to write. The input program is formulated as
follows:
%
\begin{minted}{haskell}
foldl :: ((a -o b -o a) -> a -o (List b) -o a);
newMArray :: (Int -> (MArray a -o !b) -o b);
write :: (MArray a -o (Int * a) -> MArray a);
freeze :: (MArray a -o !(Array a));
synth array :: (Int -> List (!(Int * a)) -> Array a)
    | using (foldl) | depth 3;
\end{minted}
%
And, in a small fraction of a second, matching exactly the linear definition for
\emph{array} as seen in Linear Haskell~\cite{Bernardy_2018}, \synname\ outputs:
%
\begin{minted}{haskell}
array b c = let !d = b in let !e = c in
    newMArray (!d) (\j -> freeze (foldl (!write) j e));
\end{minted}

\section{Synthesis Framework}\label{sec:formal_system}
% with Linear Types, Polymorphism and Refinements

@@

Program synthesis from linear types in combination with polymorphism, recursive
algebraic data types, and refinements, is new in the synthesis 
literature. Despite the substantial amount of research on linear logic and
proof-search upon which we found our core synthetiser, formal guidelines for the
richer types and their inherent complex problems (such as infinite recursion)
don't exist. % TODO: Developed an algorithm...?

With this section we aim to formalize as rules the novel techniques that guide
synthesis from our more expressive specifications, alongside the already well
defined rules that model the core of the synthetiser, putting together a
\emph{sound} set of inference rules that characterizes our framework for synthesis of
recursive programs from specifications with the select richer types in a linear context,
% TODO: Better?....
and describes the system in enough detail for the synthetiser to be reproducible
by a theory-driven implementation.

Even so, note that \emph{soundness} and \emph{completeness} have two very
different meanings, and our \emph{sound} set of rules only guarantees we won't
synthetise incorrect programs, but still hides a possibly infinite number of correct
proofs; and that the valid programs derivable through it reflect the somewhat
subjective trade-offs we decided on. Different choices and approaches outside
the core might lead to completely distinct synths and spaces of valid programs.

\mypara{Core Rules} The core system comprises of proof-search in sequent
calculus rules for linear logic with resource-management and focusing.
%
Andreoli's \emph{focusing} for linear logic~\cite{Andreolli
focusing} is a technique to remove inessencial non-determinism from
proof-search by structuring the application of invertible and non-invertible
inference rules. Rules are classified as \emph{left} or \emph{right},
\emph{invertible} or \emph{non-invertible} as follows:...\todo{não estou a
conseguir formular bem esta introdução}
The proof search is separated into two phases:
%
inversion ($\Uparrow$), in which we apply all invertible rules eagerly, and
focusing ($\Downarrow$), in which we decide a proposition to focus on, and then
apply non-invertible rules, staying focused until either the focus is again an
asynchronous, i.e. invertible, proposition, or the proof is complete.  For the
invertible rules, a third, ordered, context ($\Omega$) is needed to hold
propositions resulting from decomposing connectives; propositions we'll later
try to invert, moving them to the linear context ($\Delta$) when we fail to.

@@

The core language is a simply-typed linear lambda calculus with linear
functions, additive and multiplicative pairs, multiplicative unit, additive sums
and the exponential modality (to internalize unrestricted use of variables), for
which the terms ($M,N$) and types' ($\tau$) syntax is constructed with the
following grammar:
%
% TODO: Nao é um problema eu dizer que o A e B são átomos ali em cima e aqui
% usá-los como tipos gerais?
\[
\begin{tabular}{lclc}
$M,N$ & $\ ::=\ $ & $u, v$ & \emph{variables} \\
    & & $\vert\ \lambda x. M\ \vert\ M\ N$ & $\lolli$ \\
    & & $\vert\ M \with N\ \vert\ \mathsf{fst}\ M\ \vert\ \mathsf{snd}\ M$ & $\with$ \\
    & & $\vert\ M \tensor N\ \vert\ \llet{u\tensor v = M}{N}$ & $\tensor$ \\
    & & $\vert\ \star\ \vert\ \llet{\star = M}{N}$ & $\textbf{1}$ \\
    & & $\vert\ \mathsf{inl}\ M\ \vert\ \mathsf{inr}\ M\ \vert\ (\mathsf{case}\
    M\ \mathsf{of\ inl}\ u \Rightarrow N_1\ \vert\ \mathsf{inr}\ v \Rightarrow
    N_2)$ & $\oplus$ \\
    & & $\vert\ \bang M\ \vert\ \llet{\bang u = M}{N}$ & $\bang$ \\
    \\
$\tau$ & $\ ::=\ $ & $A, B, C\ \vert\ \tau \lolli \tau\ \vert\ \tau \with \tau\
    \vert\ \tau \tensor \tau\ \vert\ \textbf{1}\ \vert\ \tau \oplus \tau\ \vert\ \bang \tau$
\end{tabular}
\]
%
Putting together linear logic and linear lambda calculus through the Curry-Howard correspondence, resource
management, and focusing, we get the following core formal system (heavily inspired
by Frank Pfenning's notes on focusing for linear logic~\cite{FPnotes...}) -- in which the
first rule is read: to synthetise a program of type $A \lolli B$ while inverting
right (the $\Uparrow$ on the goal), with unrestricted context $\Gamma$, linear
context $\Delta$, and inversion context $\Omega$, synthetise a program of type $B$ with
an additional hypothesis of type $A$ named $x$ in the $\Omega$ context,
resulting in the program $M$ and output linear
context $\Delta'$ that cannot contain the added hypothesis $x{:}A$. Finally, the
resulting program is $\lambda x . M$ and the output linear context is
$\Delta'$.

@@

The decision rules are followed by either left or right focus rules:
\begin{mathpar}
    \infer*[right=($\tensor R$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow \and \Gamma ; \Delta'/\Delta'' \vdash N
    : B \Downarrow}
    {\Gamma; \Delta/\Delta'' \vdash (M \tensor N) : A \tensor B \Downarrow}
\and
    \infer*[right=($1 R$)]
    { }
    {\Gamma; \Delta/\Delta \vdash \star : \textbf{1} \Downarrow}
\and
    \infer*[right=($\oplus R_1$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inl}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\oplus R_2$)]
    {\Gamma; \Delta/\Delta' \vdash M : B \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inr}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\bang R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash M : A \Uparrow \and \Delta = \Delta'}
    {\Gamma; \Delta/\Delta \vdash \bang M : \bang A \Downarrow}
\and
    \infer*[right=($\lolli L$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C \and \Gamma;
    \Delta'/\Delta''; \cdot \vdash N : A \Uparrow}
    {\Gamma; \Delta/\Delta''; x{:}A \lolli B \Downarrow\ \vdash M\{(x\,N)/y\} : C}
\and
    \infer*[right=($\with L_1$)]
    {\Gamma; \Delta/\Delta'; y{:}A \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{fst}\ x)/y\} : C}
\and
    \infer*[right=($\with L_2$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{snd}\ x)/y\} : C}
\end{mathpar}
%
Eventually, the focus proposition will no longer be synchronous, i.e. it's
atomic or asynchronous. If we're left focused on an atomic proposition we either
instantiate the goal or fail. Otherwise the left focus is asynchronous and we
can start inverting it. If we're right focused on a proposition that isn't right
synchronous, we switch to inversion as well. Three rules model these conditions:
%
\begin{mathpar}
    \infer*[right=(init)]
    {  }
    {\Gamma; \Delta/\Delta'; x{:}A \Downarrow\ \vdash x : A}
\and
    \infer*[right=($\Downarrow L$)]
    {\Gamma; \Delta/\Delta'; A \Uparrow\ \vdash C \and A\ \textrm{not atomic and not left synchronous}}
    {\Gamma; \Delta/\Delta'; A \Downarrow\ \vdash C}
\and
    \infer*[right=($\Downarrow R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash A \Uparrow}
    {\Gamma; \Delta/\Delta' \vdash A \Downarrow}
\end{mathpar}
The rules written above together make the core of our synthetiser. Next, we'll
present new rules that align and build on top of these to synthetise recursive
programs from more expressive (richer) types.

>>>>

% General motivation?
Program synthesis is an automated or semi-automated process of deriving a
program, i.e.~generating code, from a high-level specification. Synthesis can be
seen as a means to improve programmer productivity and program correctness
(e.g. through suggestion and autocompletion).
%
Specifications can take many forms such as natural
language~\cite{chen2021evaluating},
examples~\cite{DBLP:conf/popl/FrankleOWZ16} or rich types such as
polymorphic refinement types~\cite{DBLP:conf/pldi/PolikarpovaKS16} or
graded types~\cite{DBLP:conf/lopstr/HughesO20}. Regardless of the kind
of specification, program synthesis must deal with two main
inherent sources of complexity -- search over the space of
valid programs, and interpreting user intent.

@@

The focus of our work is on type-driven synthesis where specifications take the
form of linear types.
%
% TODO: Deixar isto para mais à frente?  More concretely, we will explore the
% proof-theoretic foundations of linear types via the Curry-Howard
% correspondence with linear logic propositions, viewing program synthesis as
% proof search in linear logic.  (Said correspondence maps \emph{types} to
% \emph{propositions in a logic}, and \emph{programs} to \emph{proofs}).
%
Linear types constrain resource
usage in programs by \emph{statically} limiting the number of times certain
resources can be used during their lifetime (linear resources must be used
\emph{exactly once}).
%
They can be applied to resource-aware programming such as concurrent programming
(e.g. session types for message passing
concurrency~\cite{DBLP:journals/mscs/CairesPT16}), memory-management
(e.g.~Rust's ownership types), safely updating-in-place mutable
structures~\cite{Bernardy_2018}, enforcing protocols for external \textsc{api}s~\cite{Bernardy_2018}, to name a few.
% TODO: Repensar esta parte dos exemplos de aplicação e de como podemos
% relacionar o nosso trabalho com futuras aplicações nestes campos

\paragraph{Contributions and Outline.} Despite their long-known potential
\cite{Wadler90lineartypes,DBLP:journals/mscs/CairesPT16,Bernardy_2018}
and strong proof-theoretic
foundations
\cite{10.1093/logcom/2.3.297,DBLP:conf/cade/ChaudhuriP05,DBLP:journals/tcs/CervesatoHP00},
synthesis with linear types combined with other advanced typing features
has generally been overlooked in the literature.  In this work we present a framework for synthesis
with linear types extended with recursive algebraic data types, parametric
polymorphism and refinements.
% TODO: extensible? proving feasability of... ? showing potential for ... ?
We first introduce linear types as specifications
and outline the synthesis process, leveraging linearity, by example
(\S~\ref{sec:overview}). We then present the formal system driving the
synthesis (\S~\ref{sec:formal_system}) and describe the architecture
of our framework named \synname, examining technical details and key
implementation challenges (\S~\ref{sec:architecture}). Finally,
we evaluate our work through expressiveness and performance benchmarks
(\S~\ref{sec:evaluation}) and discuss related work (\S~\ref{sec:related}).
%TODO: And appendice with...

@@

The \synname\ synthetiser combines linear types with recursion,
parametric polymorphism, recursive algebraic data types, and
refinement types. The synthetiser is built on top of a system of proof-search for
linear logic. Proof search relates to program synthesis via the Curry-Howard
correspondence~\cite{}, which states that \emph{propositions in a logic} are
\emph{types}, and \emph{proofs} of those propositions are well-typed
\emph{programs} -- finding a proof of a proposition is finding a program with
that type.

Linear types make for more precise specifications than simple types
because information on which resources must necessarily be used is
encoded in the type. For instance the type $\mathsf{Int} \lolli
\mathsf{Int} \lolli \mathsf{Int}$ denotes a function that \emph{must}
use its two integer arguments to produce an integer.
Their preciseness also affects the search space:
all programs where a linear resource is used non-linearly (i.e. not
\emph{exactly once}) are ill-typed. With linearity built into the
synthesis process, usage of a linear proposition more than once is not
considered, and unused propositions are identified during synthesis,
constraining the space of valid programs.  % TODO: Reword?

The core of the synthesis is a \emph{sound} and \emph{complete} system
consisting of \emph{bottom-up} proof-search in propositional linear
logic based on \emph{focusing}~\cite{}. Our approach, being grounded
by propositions-as-types, ensures that all synthetised programs
(i.e.~proofs) are well-typed \emph{by construction} (i.e.~if the
synthesis procedure produces a program, then the program intrinsically
satisfies its specification). Moreover, we can leverage the modularity
of the proof-search based approach along two axes: first, since proof
search need not construct only closed proofs, we can effectively
synthesize program sub-expressions (i.e.~synthesis based on typed
holes); secondly, the framework is amenable to extensions to the core
propositional language, allowing for the introduction of a richer type
structure while preserving the correctness of programs by
construction.

The \synname\ programming language is naturally developed alongside
the synthetiser. Synthesis goals are inserted in the program by use of
a \emph{synth} keyword or mark, indicating a program should be
generated for a given type. Additionally, it opens the possibility of
synthesis within a context, that is, with knowledge of other functions and
structures defined in the same program.

\paragraph{Core linear calculus.} Initially, we synthesize from
specifications that only use literals and canonical linear types ($\lolli, \with, \tensor, 1,
\oplus, \bang$), such as linear logic theorems. Namely,
the specification $(A \lolli B \lolli C) \lolli (A \tensor B) \lolli C$ produces
the function \mintinline{haskell}{\a -> \b -> let c*d = b in a c d},
which takes an argument \mintinline{haskell}{a}, a
\emph{curried} linear function of type $A \lolli B \lolli C$,  and an
argument \mintinline{haskell}{b}, a \emph{linear
  pair} of an $A$ and a $B$ and produces a value of type $C$ by
deconstructing the pair and applying \mintinline{haskell}{a} to the
corresponding elements of the pair (i.e.,~\emph{uncurries} the
function \mintinline{haskell}{a}).

\paragraph{Beyond propositional logic.} To be able to synthesize more
interesting programs -- and thus, empirically prove the feasability of
more relevant synthesis in a linear context, we extend the syntax and
type system with recursiveness, parametric polymorphism, algebraic
data types (ADTs) and type refinements.

@@

\todo[inline]{Tens de explicar o que significa o $\bang$, see above.}
\paragraph{ADTs.} We reiterate the expressiveness of linear types: by requiring some types to be
used linearly (consumed), we can assure, e.g., the deconstruction of an
ADT. As such, the specification $\bang(a \lolli b) \lolli List\ a \lolli
List\ b$ isn't satisfied by the program \mintinline{haskell}{(\x -> \y -> Nil)}.
To synthetise a valid program from it, \mintinline{haskell}{(List a)} has to be
deconstructed, and its constructor arguments used. \synname\ would output the following
function (\emph{map}) without relying on any additional information:
% TODO: Aqui não estou bem a falar de polymorphism mas parece estranho não por
% visto que falo dos outros temas? se calhar faço um à parte?
%
\begin{minted}{haskell}
map c d = let !e = c in
    case d of
          Nil  -> Nil
        | Cons f -> let g*h = f in Cons (e g, map (!e) h);
\end{minted}

\todo[inline]{Tens de dizer o que significa o refinamento concreto e o
que e a funcao mult e add}
\paragraph{Refinements.} Refinements are logical predicates
which must hold for any element that inhabits a given type. The \synname\
language supports simple refinements on integers with arithmetic
expressions in the predicates. They assert the robustness of our main
framework, i.e., how it is amenable to significant additions without
necessarily interfering with the overall approach.
%
As an example, from the specification $x \{Int\} \lolli y \{Int\} \lolli z
\{Int\} \lolli k \{Int\ \vert\ x + k = y * z\}$, the
program\ \mintinline{haskell}{(fn a b c = add (mult (sub 0 1) a) (mult c b))} is
synthetised.

\paragraph{Guiding the Search.} Specifications can be additionally
augmented with three keywords: ``using'', ``depth'' and
``assert''. Through them, we can fine-tune the synthesis process and,
respectively, force certain functions to be present in the synthesis
outcome, allow a \emph{deeper} search in the valid programs space, and
require given predicates to hold true in the program after the
synthesis is complete.

Taken directly from the Linear Haskell paper~\cite{Bernardy_2018}, we present a
more intricate example of synthesis: given the linear type signatures for array
primitives (\emph{newMArray} passes a new mutable array to a function that uses
it linearly, \emph{write} takes a mutable array and writes a pair to it,
\emph{freeze} consumes a mutable array and produces an imutable array,
\emph{foldl} has the default definition) we synthesize a function
``array'' that provides an immutable array given a list of pairs to write. The input program is formulated as
follows:
%
\begin{minted}{haskell}
foldl :: ((a -o b -o a) -> a -o (List b) -o a);
newMArray :: (Int -> (MArray a -o !b) -o b);
write :: (MArray a -o (Int * a) -> MArray a);
freeze :: (MArray a -o !(Array a));
synth array :: (Int -> List (!(Int * a)) -> Array a)
    | using (foldl) | depth 3;
\end{minted}
%
Matching exactly the linear definition for
\emph{array} from~\cite{Bernardy_2018}, \synname\
outputs (in a fraction of a second):
%
\begin{minted}{haskell}
array b c = let !d = b in let !e = c in
    newMArray (!d) (\j -> freeze (foldl (!write) j e));
\end{minted}

\section{The \synname\ Synthesis Framework}\label{sec:formal_system}
% with Linear Types, Polymorphism and Refinements

@@

Program synthesis from linear types with polymorphism, recursive
algebraic data types, and refinements, is essentially new in the synthesis 
literature. Despite the substantial amount of research on linear logic and
proof-search upon which we base our core synthesizer, formal guidelines for
richer types and their intrinsic challenges (such as infinite
recursion) must be developed.

In this section we formalize  the techniques that guide
synthesis from our more expressive specifications, alongside the already well
defined rules that model the core of the synthetiser, putting together a
\emph{sound} set of inference rules that characterizes our framework
for linear synthesis of
recursive programs from specifications with the select richer types,
% TODO: Better?....
and describes the system in enough detail for the synthetiser to be reproducible
by a theory-driven implementation.

We note that a \emph{sound} set of rules guarantees we cannot synthesize
incorrect programs; and that the valid programs derivable through them
reflect the somewhat subjective trade-offs we committed to. Different
choices and approaches outside the core might lead to completely
distinct synths and spaces of valid programs.

\todo[inline]{Indicar o judgment ``declarativo'' e depois os
  judgments em focusing}

\mypara{Core Rules} The core system comprises of proof search for
(intuitionistic) linear logic sequent calculus proofs, based on a
system of resource management and focusing.
%
In intuitionistic sequent calculi, each connective has a so-called
\emph{left} and a \emph{right} rule, which effectively define how to
decompose an ambient assumption of a given proposition and how to
prove a certain proposition is true, respectively. 
Andreoli's \emph{focusing} for linear logic~\cite{Andreolli focusing}
is a technique to remove non-essential nondeterminism from
proof-search by structuring the application of so-called invertible
and non-invertible inference rules. Andreoli observed that the
connectives of linear logic can be divided into two categories, dubbed
synchronous and asynchronous. Synchronous connectives are those whose
right rules are \emph{invertible}, i.e. they can be applied eagerly
during proof search without losing provability and so the order in
which these rules are applied is irrelevant, and whose left rules
are not invertible. Asynchronous connectives are dual. \todo{Say which
  are which}
%
Given this separation, focusing divides proof search into two
phases: %
the inversion phase ($\Uparrow$), in which we apply \emph{all} invertible rules
eagerly, and the focusing phase  ($\Downarrow$), in which we decide a proposition
to focus on, and then apply non-invertible rules, staying \emph{in focus}
until we reach an asynchronous, i.e. invertible,
proposition or the proof is complete.

\todo[inline]{Isto so faz sentido olhando para os judgments}
For the invertible rules, a
third, ordered, context ($\Omega$) is needed to hold propositions
resulting from decomposing connectives; propositions we'll later try
to invert, moving them to the linear context ($\Delta$) when we fail
to.

@@

\todo[inline]{A frente de cada ``descricao'', escreve o tipo pois
  ajuda o leitor}
The core language is a simply-typed linear lambda calculus with linear
functions, additive and multiplicative pairs, multiplicative unit, additive sums
and the exponential modality (to internalize unrestricted use of variables), for
which the terms ($M,N$) and types' ($\tau$) syntax is constructed with the
following grammar:
%
% TODO: Nao é um problema eu dizer que o A e B são átomos ali em cima e aqui
% usá-los como tipos gerais? NAO :)
\[
\begin{tabular}{lclc}
$M,N$ & $\ ::=\ $ & $u, v$ & \emph{variables} \\
    & & $\vert\ \lambda x. M\ \vert\ M\ N$ & $\lolli$ \\
    & & $\vert\ M \with N\ \vert\ \mathsf{fst}\ M\ \vert\ \mathsf{snd}\ M$ & $\with$ \\
    & & $\vert\ M \tensor N\ \vert\ \llet{u\tensor v = M}{N}$ & $\tensor$ \\
    & & $\vert\ \star\ \vert\ \llet{\star = M}{N}$ & $\textbf{1}$ \\
    & & $\vert\ \mathsf{inl}\ M\ \vert\ \mathsf{inr}\ M\ \vert\ (\mathsf{case}\
    M\ \mathsf{of\ inl}\ u \Rightarrow N_1\ \vert\ \mathsf{inr}\ v \Rightarrow
    N_2)$ & $\oplus$ \\
    & & $\vert\ \bang M\ \vert\ \llet{\bang u = M}{N}$ & $\bang$ \\
    \\
$\tau$ & $\ ::=\ $ & $A, B, C\ \vert\ \tau \lolli \tau\ \vert\ \tau \with \tau\
    \vert\ \tau \tensor \tau\ \vert\ \textbf{1}\ \vert\ \tau \oplus \tau\ \vert\ \bang \tau$
\end{tabular}
\]
%
Putting together linear logic and linear lambda calculus through the Curry-Howard correspondence, resource
management, and focusing, we get the following core formal system (heavily inspired
by Frank Pfenning's notes on focusing for linear logic~\cite{FPnotes...}) -- in which the
first rule \todo{qual ``first''? referencia por nome} is read: to synthesize a program of type $A \lolli B$ while inverting
right (the $\Uparrow$ on the goal), with unrestricted context $\Gamma$, linear
context $\Delta$, and inversion context $\Omega$, synthetise a program of type $B$ with
an additional hypothesis of type $A$ named $x$ in the $\Omega$ context,
resulting in the program $M$ and output linear
context $\Delta'$ that cannot contain the added hypothesis $x{:}A$. Finally, the
resulting program is $\lambda x . M$ and the output linear context is
$\Delta'$.

@@

\todo[inline]{Aqui tens de explicar uma regra pelo menos}
The decision rules are followed by either left or right focus rules:
\begin{mathpar}
    \infer*[right=($\tensor R$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow \and \Gamma ; \Delta'/\Delta'' \vdash N
    : B \Downarrow}
    {\Gamma; \Delta/\Delta'' \vdash (M \tensor N) : A \tensor B \Downarrow}
\and
    \infer*[right=($1 R$)]
    { }
    {\Gamma; \Delta/\Delta \vdash \star : \textbf{1} \Downarrow}
\and
    \infer*[right=($\oplus R_1$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inl}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\oplus R_2$)]
    {\Gamma; \Delta/\Delta' \vdash M : B \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inr}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\bang R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash M : A \Uparrow \and \Delta = \Delta'}
    {\Gamma; \Delta/\Delta \vdash \bang M : \bang A \Downarrow}
\and
    \infer*[right=($\lolli L$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C \and \Gamma;
    \Delta'/\Delta''; \cdot \vdash N : A \Uparrow}
    {\Gamma; \Delta/\Delta''; x{:}A \lolli B \Downarrow\ \vdash M\{(x\,N)/y\} : C}
\and
    \infer*[right=($\with L_1$)]
    {\Gamma; \Delta/\Delta'; y{:}A \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{fst}\ x)/y\} : C}
\and
    \infer*[right=($\with L_2$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{snd}\ x)/y\} : C}
\end{mathpar}
%
Eventually, the focus proposition will no longer be synchronous, i.e. it's
atomic or asynchronous. If we're left focused on an atomic proposition we either
instantiate the goal or fail. Otherwise the left focus is asynchronous and we
can start inverting it. If we're right focused on a proposition that isn't right
synchronous, we switch to inversion as well. Three rules model these conditions:
%
\begin{mathpar}
    \infer*[right=(init)]
    {  }
    {\Gamma; \Delta/\Delta'; x{:}A \Downarrow\ \vdash x : A}
\and
    \infer*[right=($\Downarrow L$)]
    {\Gamma; \Delta/\Delta'; A \Uparrow\ \vdash C \and A\ \textrm{not atomic and not left synchronous}}
    {\Gamma; \Delta/\Delta'; A \Downarrow\ \vdash C}
\and
    \infer*[right=($\Downarrow R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash A \Uparrow}
    {\Gamma; \Delta/\Delta' \vdash A \Downarrow}
\end{mathpar}
The rules written above together make the core of our synthetiser. Next, we'll
present new rules that align and build on top of these to synthetise recursive
programs from more expressive (richer) types.

\mypara{Algebraic Data Types} In its simplest form, an algebraic data type (ADT)
is a tagged sum of any type, i.e. a named type that can be instantiated by one of many
tags (or constructors) that take some value of  a fixed type, which might
be, e.g., a product type ($A \tensor B$), or unit ($1$), in practice allowing
for constructors with an arbitrary number of parameters. In the \synname\
language, the programmer can define custom ADTs; as an example, we show the
definition of an ADT which holds zero, one, or two
values of type $A$, using the syntax: \mintinline{haskell}{data Container
= None 1 | One A | Two (A * A)}. The grammar is extended as:
\[
\begin{tabular}{lclc}
    $M,N$ & $\ ::=\ $ & $\dots\ \vert\ \emph{Cons}_n\ M\ \vert\ (\mathsf{case}\
    M\ \mathsf{of}\ \dots\ \vert\ \emph{Cons}_n\ u \Rightarrow N)$ \\
    $\tau$ & $\ ::=\ $ & $\dots\ \vert\ \langle\emph{name}\rangle$ \\
    $ADT$ & $\ ::=\ $ & (\textsf{data} $\langle\emph{name}\rangle$\ $=$\ $\emph{Cons}_1\ \tau$\
$\vert$\ $\dots$\ $\vert$\ $\emph{Cons}_n\ \tau$) \\ \end{tabular}
\]
%TODO: Como definir gramática dos ADTs?
